---
title: "correlations_exercise"
author: "Alex Faidiga"
date: "May 4, 2020"
output: html_document
---
You can do this on your own data, or on included data here.

##Continuous data

```{r setup, eval = TRUE}
library(geiger)
library(ape)
tree.primates <- read.tree(text="((((Homo:0.21,Pongo:0.21):0.28,Macaca:0.49):0.13,Ateles:0.62):0.38,Galago:1.00);") #using examples from ape ?pic
plot(tree.primates)
X <- c(4.09434, 3.61092, 2.37024, 2.02815, 1.46968) 
Y <- c(4.74493, 3.33220, 3.36730, 2.89037, 2.30259) 
names(X) <- names(Y) <- c("Homo", "Pongo", "Macaca", "Ateles", "Galago")
pic.X <- pic(X, tree.primates)
pic.Y <- pic(Y, tree.primates)

#Now, positivitize the contrasts and do a regression through the origin.
PIC1 <- lm(pic.Y ~ pic.X)
summary(PIC1)

plot(pic.X, pic.Y)
abline(PIC1$coefficients[1], PIC1$coefficients[2])

```

##Discrete data

```{r, eval = TRUE}
require("corHMM")
?corHMM
data(primates)
ls()
print(primates)
require(phytools)
```

Just to make this a better dataset for our exercise, I’m going to change some of the states (I want to have all four trait combinations present). For actual analyses, of course, DO NOT INVENT YOUR DATA.

First, a review of discrete state models:

```{r, eval = TRUE}
primates$trait[which(grepl("Hylobates",primates$trait[,1])),2]<-1

trait1<-primates$trait[,2]
names(trait1)<-primates$trait[,1]
primates$tree <- ape::multi2di(primates$tree)
plotSimmap(make.simmap(primates$tree, trait1), pts=FALSE, fsize=0.8)
rate.mat.er<-rate.mat.maker(rate.cat=1, hrm=FALSE, ntraits=1, nstates=2, model="ER")
print(rate.mat.er)
```

####What does this matrix mean?
This sets up the rate matrix for transitions among states; in this case, r1->2 = r2->1.

```{r, eval = TRUE}
pp.er<-corHMM(primates$tree,primates$trait[,c(1,2)],rate.cat=1,rate.mat=rate.mat.er,node.states="marginal")
print(pp.er)
```

####What do these results mean?
That the rate of transition from state 1 to 2 or 2 to 1 is 0.0107. 

```{r, eval = TRUE}
rate.mat.ard<-rate.mat.maker(rate.cat=1, hrm=FALSE, ntraits=1, nstates=2, model="ARD")
print(rate.mat.ard)
```

####And these?
This model allows the two rates to be different. 

```{r, eval = TRUE}
pp.ard<-corHMM(primates$tree,primates$trait[,c(1,2)],rate.cat=1,rate.mat=rate.mat.ard,node.states="marginal")
print(pp.ard)
```

####Which model is better?
The first model where the two rates are the same is better because it has a lower AICc valule. 

####Now let’s look at multiple traits. This is a matrix with four states:

```{r, eval = TRUE}
rate.mat.er.4state<-rate.mat.maker(rate.cat=1, hrm=FALSE, ntraits=1, nstates=4, model="ER")
print(rate.mat.er.4state)

#Convert the two binary traits into a single four character state

fourstate.trait<-rep(NA,Ntip(primates$tree))
for(i in sequence(Ntip(primates$tree))) {
    if(primates$trait[i,2]==0 && primates$trait[i,3]==0) {
        fourstate.trait[i]<-0
    }
    if(primates$trait[i,2]==0 && primates$trait[i,3]==1) {
        fourstate.trait[i]<-1
    }
    if(primates$trait[i,2]==1 && primates$trait[i,3]==0) {
        fourstate.trait[i]<-2
    }
    if(primates$trait[i,2]==1 && primates$trait[i,3]==1) {
        fourstate.trait[i]<-3
    }
}
fourstate.data<-data.frame(Genus_sp=primates$trait[,1], T1=fourstate.trait)

print(rayDISC(primates$tree, fourstate.data, ntraits=1, model="ER", node.states="marginal"))

rate.mat.ard.4state<-rate.mat.maker(rate.cat=1, hrm=FALSE, ntraits=1, nstates=4, model="ARD")

print(rate.mat.ard.4state)

print(rayDISC(primates$tree, fourstate.data, ntraits=1, rate.mat=rate.mat.ard.4state, node.states="marginal", model="ARD"))
```
Now let’s make the equivalent of a GTR matrix:

```{r, eval = FALSE}
rate.mat.gtr.4state<-rate.mat.ard.4state
rate.mat.gtr.4state<-rate.par.eq(rate.mat.gtr.4state, c(1,4))
rate.mat.gtr.4state<-rate.par.eq(rate.mat.gtr.4state, c(2,6))
rate.mat.gtr.4state<-rate.par.eq(rate.mat.gtr.4state, c(3,8))
rate.mat.gtr.4state<-rate.par.eq(rate.mat.gtr.4state, c(4,6))
rate.mat.gtr.4state<-rate.par.eq(rate.mat.gtr.4state, c(5,7))
rate.mat.gtr.4state<-rate.par.eq(rate.mat.gtr.4state, c(6,7))
print(rate.mat.gtr.4state)

print(rayDISC(primates$tree, fourstate.data, ntraits=1, rate.mat= rate.mat.gtr.4state, node.states="marginal", model="ARD"))
```

Now make a model like Pagel 1994

```{r, eval = FALSE}
print(rate.mat.maker(rate.cat=1, hrm=FALSE, ntraits=2, nstates=2, model="ARD"))
rate.mat.pag94<-rate.par.drop(rate.mat.ard.4state, drop.par=c(3,5,8,10))

print(rayDISC(primates$tree, fourstate.data, ntraits=1, rate.mat= rate.mat.pag94, node.states="marginal", model="ARD"))
```

Now that you have some introduction, there are two routes:

##Route 1

Construct a model to test if state 1 can never be lost.

```{r, eval = FALSE}
rate.mat.ard.4state<-rate.mat.maker(rate.cat=1, hrm=FALSE, ntraits=1, nstates=4, model="ARD")

rate.mat.ard.noloss1<-rate.par.drop(rate.mat.ard.4state, drop.par=c(1:3, 5:6,8:9))

print(rayDISC(primates$tree, fourstate.data, ntraits=1, rate.mat= rate.mat.ard.noloss1, node.states="marginal", model="ARD"))
```

Create and use a model to see if transitions from 00 go to 11 only via 01.

```{r, eval = FALSE}
rate.mat.ard.4state<-rate.mat.maker(rate.cat=1, hrm=FALSE, ntraits=1, nstates=4, model="ARD")

rate.mat.ard.onlyvia01<-rate.par.drop(rate.mat.ard.4state, drop.par=c(7,10))

print(rayDISC(primates$tree, fourstate.data, ntraits=1, rate.mat= rate.mat.ard.onlyvia01, node.states="marginal", model="ARD"))
plot(primates$tree)
```

Experiment with the effects of frequencies at the root.

