---
title: "R Markdown Practice"
author: "Alex Faidiga"
date: "January 20, 2020"
output: html_document
---
## Code
	 
Pieces of code are called chunks. 
	 
```{r eval = TRUE}
	print("This is a chunk.")
```
	 
To suppress the code but still show the output, use echo = FALSE.
	 
```{r echo = FALSE}
	print("This is another chunk.")
```
	 
To suppress the output but still show the code, use results = "hide".
	 
```{r results = "hide"}
	print("This is yet another chunk.")
```
	 
You can include plots:
	 
```{r eval = TRUE}
	plot(rnorm(50))
```
	 
Code within text: 
	 
The square root of 1421 is `r sqrt(1421) `.
	 
## Formatting text 
	 
You can make the main text of the document *italicized* and **bold**, or create headers:
	 
# This
## Is 
### Rmarkdown
#### Practice
	 
Create a break in the document using three stars:
	 
***
	 
Add quotes with a single greater-than symbol (>):
	 
>Hello world 
	
Make lists with bullet points using * and +:
	 
* Groceries
	  + Milk
	  + Eggs
	  + Avocados
	  
Make tables with lines! 
	 
Plant   | Ranking
--------|--------
trees   | 1
flowers | 2
moss    | 3
	 
	


