---
title: "Discrete characters exercise"
author: "Alex Faidiga"
date: "February 17, 2020"
output: html_document
---
```{r packages, include = FALSE}
setwd("~/Desktop/phylometh")
library(geiger)
library(rotl)
library(ape)
library(phytools)
library(phangorn)
library(corHMM)
```
 
Loading in data from Smith and Brown's (2017) huge tree of angiosperms, downloaded from https://github.com/FePhyFoFum/big_seed_plant_trees/releases. Both of these trees are built with GenBank data, but they differ in what tree is used as a "backbone". The GBOTB tree uses Open Tree of Life data, and the GBMB uses data from Mallagón et al. (2015).

```{r viola tree}
GBOTB_tree <- read.tree("GBOTB.tre")
GBMB_tree <- read.tree("GBMB.tre")
ALLMB_tree <- read.tree("ALLMB.tre")
```

I'll create a function that isolates a clade from the giant tree by identifying the node with the MRCA of two species provided to the function:

```{r visualize trees}

clade_finder <- function(species1, species2, tree){
  viola_node<-fastMRCA(tree,species1,species2)   #identifies the most recent node where these two species shared a common ancestor 
  viola_clade<-extract.clade(tree, viola_node) #extracts and creates a new tree for a clade based on the MRCA node identified  
plot.phylo(viola_clade, cex = 0.3, type = "fan")
print(Ntip(viola_clade))
}

clade_finder("Viola_pubescens", "Viola_rostrata", GBOTB_tree)
clade_finder("Viola_pubescens", "Viola_rostrata", GBMB_tree)

```

Now I have a clade of violets so I know the data are ok, but it's not necessarily a meaningful clade. Between the two large trees, it looks like the clades I extracted have the same topology and number of tips. 

Now, I'll extract all violets from the large tree using genbank data (GBMB_tree and GBOTB_tree) and plot them. 

```{r viola tree attempt 1}
# length(GBMB_tree$tip.label)
# length(GBOTB_tree$tip.label) #this tree has 7 more tips

all_tips_GBMB <- GBMB_tree$tip.label
all_tips_GBOTB <- GBOTB_tree$tip.label

viola_tips_GBMB <- all_tips_GBMB[grepl("Viola_", fixed = TRUE, GBMB_tree$tip.label)]
viola_tips_GBOTB <- all_tips_GBOTB[grepl("Viola_", fixed = TRUE, GBOTB_tree$tip.label)]

# length(viola_tips_GBMB) 
# length(viola_tips_GBOTB) #same for both; 173 viola spp.

viola_node<-getMRCA(GBMB_tree, viola_tips_GBMB)   
viola_tree<-extract.clade(GBMB_tree, viola_node)

plot(viola_tree, cex = 0.5)

#write to PDF and open using preview
pdf(file="~/Desktop/phylometh/viola_tree1.pdf", width=20, height=20)

  plot(viola_tree, cex = 0.5, type = "fan")

system("open -a Preview.app ~/Desktop/phylometh/viola_tree1.pdf")
dev.off()
```

Another way to extract species of interest and build a tree if they are paraphyletic:

```{r viola tree attempt 2}
viola_tree2 <- drop.tip(GBMB_tree, GBMB_tree$tip.label[!(GBMB_tree$tip.label %in% viola_tips_GBMB)])

pdf(file="~/Desktop/viola_tree2.pdf", width=20, height=20)

  plot(viola_tree2, type = "fan", cex = 1)

dev.off()
system("open ~/Desktop/viola_tree2.pdf")
```

###It’s critically important to LOOK at what you have. Are there weird values? Has the match between taxa and state gone correctly? Do you think you have binary data, but there’s actually only state 1? Especially as data sets grow (yay), and are assembled using scripts rather than by error-prone, non-reproducable hands (double yay), scientists are increasingly less likely to deeply look at our data. That’s bad – don’t be that person.

```{r data visualization function}
VisualizeData <- function(phy, data) {
  fmode<-as.factor(setNames(data[,1],rownames(data)))
  trait1 <- unique(data[,1])[1]
  trait2 <- unique(data[,1])[2]
  trait3 <- unique(data[,1])[3]
  dotTree(phy, fmode, colors=setNames(c("blue","red", "green"),
    c(paste(trait1), paste(trait2), paste(trait3))),ftype="i", fsize = 0.3) #plot the discrete character on the phylogeny

}

#Create random data (in data frame form to work with this function) for viola since I don't have any:
species_list <- viola_tree$tip.label
viola_binary_trait <- rbinom(n=173, size=2, prob=0.3)
#viola_cleistogamy <- strrep("CH/CL", "CH only", "unknown", times = 1:3)
viola_data <- data.frame(species_list, viola_binary_trait, row.names = 1)

pdf(file="~/Desktop/viola_tree_traits.pdf", width=20, height=20)

VisualizeData(viola_tree, viola_data)

dev.off()
system("open ~/Desktop/viola_tree_traits.pdf")

```

Test with another data set:

```{r}
eel_tree<-read.tree("elopomorph.txt")
eel_data<-read.csv("elopomorph.csv",row.names=1)

VisualizeData(eel_tree, eel_data)
#yay! it works!

```

**NOTE: this function works for data frames where each row name is a species and there is a column for discrete trait data, where each species has a given value for that trait. Trait value can be binary or a string.**

###First, let’s use parsimony to look at ancestral states:

```{r eval = FALSE}
cleaned.discrete.phyDat <- phangorn::phyDat(viola_data, type="USER", levels = c(1,0)) #phyDat is a data format used by phangorn
anc.p <- phangorn::ancestral.pars(tree, cleaned.discrete.phyDat)
plotAnc(tree = viola_tree, data = anc.p, i = 1, cex.pie = 0.2, cex = 0.3, type = "fan")

pdf(file="~/Desktop/anc.states.parsimony.pdf", width=20, height=20)

plotAnc(tree = viola_tree, data = anc.p, i = 1, cex.pie = 0.2, cex = 0.3, type = "fan")

dev.off()
system("open -a Preview.app ~/Desktop/ancestral.states.pdf")
```

###Do you see uncertainty? What does it mean?

It seems like uncertainty increases as it gets closer to the tips. This sort of makes sense since my data is just random numbers from 0-2; the data are not very clustered at the tips, so ancestral states one node below the tips are likely not very certain. As we move further down the tree, the estimates for ancestral states likely become more certain because there are more data used for calculating the state at each node.

###Now, plot the likelihood estimates.

```{r eval = FALSE}
anc.ml <- ancestral.pml(pml(tree = viola_tree, data = cleaned.discrete.phyDat), type="ml")

pdf(file="~/Desktop/anc.states.ML.pdf", width=20, height=20)

plotAnc(tree = viola_tree, data = anc.ml, i = 1, cex.pie = 0.2, cex = 0.7, type = "fan")

dev.off()
system("open -a Preview.app ~/Desktop/anc.states.ML.pdf")
```

###How does this differ from parsimony?
The uncertainty in the estimates seems much greater when using ML.

###Why does it differ from parsimony?
Parsimony is estimating the ancestral states based on the least amount of changes possible, while ML is finding the ancestral states that maximize the likelihood of the data on the tree. 

###What does uncertainty mean?
Uncertainty in this case has to do with our uncertainty in the character state at a given node. 

###Now, to the biological questions. For many of these, corHMM will be a useful package. Do the following analyses:

####How can you estimate transition rates between states? Do it.

```{r}

#corHMM only works for binary traits, so I'll simulate data again omitting the 2:
species_list <- viola_tree$tip.label
viola_binary_trait2 <- rbinom(n=173, size=1, prob=0.2)

viola_data2 <- matrix(c(species_list, viola_binary_trait2), nrow=173, ncol=2, byrow=F)

cor.rates <- corHMM(phy = viola_tree, data = viola_data2, node.states = "marginal", rate.cat = 2)
cor.rates$states
cor.rates$solution

```
Things I am confused about:
1. what are r0 and r1? Like why are there secondary rates underlying each character state?
2. what units are the rates in that are returned in the matrix?


####How could you examine if transition rates are equal?
The "solution" output from corHMM tells you the transition rates.

####Think about the Lewis (2001) MKV model. Are your traits all variable? Will using this make sense for your data? Try using it. Do results change? How could you test order of state evolution?


